---
title: "My title"
subtitle: "My subtitle if needed"
author: "Shrey Sati"
thanks: "Code and data are available at: [https://github.com/UpCatch/Toronto_Tick_Surveillance_Data](https://github.com/UpCatch/Toronto_Tick_Surveillance_Data)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false

library(tidyverse)
library(here)
library(sf)
library(tinytable)

# Load the dataset
analysis_data <- read_csv(here("data", "02-analysis_data", "analysis_data.csv"))
```


# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....






# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@blacklegged_tick_data].... Following @tellingstories, we consider...

Overview text

## Measurement
	
Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

```{r}
#| label: tbl-annual-tick-summary
#| tbl-cap: "The share of collected adult and nymph ticks that tested positive rose from under 6% before 2016 to between 29% and 43% from 2016 onward. No surveillance took place in 2020–2022."
#| echo: false
#| warning: false
#| message: false

# Summarize the data by year
summary_table <- analysis_data |> 
  group_by(year) |> 
  summarize(
    `Parks surveyed` = n(),
    `Parks with ticks` = sum(total_blts > 0, na.rm = TRUE),
    `Adults and nymphs` = sum(blt_adults_and_nymphs, na.rm = TRUE),
    `Larvae` = sum(blt_larvae, na.rm = TRUE),
    `Tested positive` = sum(num_positive, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    `Positive (%)` = (`Tested positive` / `Adults and nymphs`) * 100
  ) |> 
  rename(Year = year)

# Generate the tinytable with formatting
tt(summary_table, width = 1) |> 
  format_tt(j = 2:6, digits = 0, num_mark_big = ",") |> 
  # num_fmt = "decimal" gives one decimal place (the default rounds to significant figures)
  format_tt(j = 7, digits = 1, num_fmt = "decimal", num_zero = TRUE) |> 
  # Escape the % in the column name so LaTeX prints it instead of treating it as a comment
  format_tt(escape = TRUE) |> 
  style_tt(j = 2:7, align = "r")
```

```{r}
#| label: fig-tick-surveillance
#| fig-cap: "Distribution of total blacklegged ticks collected per park across surveillance years in Toronto."
#| echo: false
#| warning: false
#| message: false

# Convert year to factor with full levels from 2013 to 2023 to display gap years
full_years <- as.character(2013:2023)

plot_data <- analysis_data |> 
  mutate(year = factor(as.character(year), levels = full_years))

# Plot raw points with horizontal jitter
set.seed(853)

ggplot(plot_data, aes(x = year, y = total_blts)) +
  geom_jitter(
    width = 0.35,
    height = 0,
    alpha = 0.4,
    size = 1.5,
    color = "darkred"
  ) +
  scale_x_discrete(drop = FALSE) +
  labs(
    x = "Surveillance Year",
    y = "Total Blacklegged Ticks (BLTs)",
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank()
  )
```

```{r}
#| label: fig-annual-total-ticks
#| fig-cap: "Total adult and nymph blacklegged ticks collected per year across all surveyed parks in Toronto."
#| echo: false
#| warning: false
#| message: false

# Aggregate total ticks by year and retain all years 2013-2023
annual_ticks <- analysis_data |> 
  group_by(year) |> 
  summarize(sum_blts = sum(total_blts, na.rm = TRUE), .groups = "drop") |> 
  complete(
    year = 2013:2023,
    fill = list(sum_blts = NA)
  ) |> 
  mutate(year = factor(year, levels = 2013:2023))

# Plot bar chart
ggplot(annual_ticks, aes(x = year, y = sum_blts)) +
  geom_col(
    fill = "#2c7bb6",
    width = 0.65,
    na.rm = TRUE
  ) +
  geom_text(
    aes(label = ifelse(is.na(sum_blts), "", scales::comma(sum_blts))),
    vjust = -0.5,
    size = 3.3,
    color = "grey20"
  ) +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.12)),
    labels = scales::comma
  ) +
  labs(
    x = "Year",
    y = "Total Blacklegged Ticks (Adults & Nymphs)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line.x = element_line(color = "grey60", linewidth = 0.4)
  )
```

```{r}
#| label: fig-positivity-rate
#| fig-cap: "From 2016 onward, between 29% and 43% of the blacklegged ticks collected in Toronto parks each year tested positive. The number under each year (n) is the count of adult and nymph ticks collected that year. 2013 and 2014 are omitted because only one and three ticks were collected, too few for a meaningful rate. No surveillance took place in 2020–2022 (shaded)."
#| echo: false
#| warning: false
#| message: false

# Calculate annual positivity rates and complete timeline for 2015-2023
positivity_data <- analysis_data |> 
  filter(year >= 2015) |> 
  group_by(year) |> 
  summarize(
    total_positive = sum(num_positive, na.rm = TRUE),
    total_tested = sum(blt_adults_and_nymphs, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    pct_positive = (total_positive / total_tested) * 100
  ) |> 
  complete(
    year = 2015:2023
  ) |> 
  mutate(
    year_label = if_else(
      is.na(total_tested),
      as.character(year),
      paste0(year, "\nn = ", total_tested)
    ),
    # Keep the years in order on the axis
    year_label = factor(year_label, levels = year_label),
    # Separate groups so the line does not bridge the 2020-2022 gap
    period = if_else(year < 2020, "before", "after")
  )

# Plot line chart
ggplot(positivity_data, aes(x = year_label, y = pct_positive, group = period)) +
  # Declare discrete x-axis
  scale_x_discrete(drop = FALSE) +
  # Shade years 2020-2022 with no surveillance 
  annotate(
    "rect",
    xmin = 5.5, xmax = 8.5, ymin = -Inf, ymax = Inf,
    fill = "grey92"
  ) +
  annotate(
    "text",
    x = 7, y = 25, label = "No surveillance",
    colour = "grey40", size = 3.5
  ) +
  geom_line(
    color = "#8B0000",
    linewidth = 0.9,
    na.rm = TRUE
  ) +
  geom_point(
    color = "#8B0000",
    size = 2.8,
    na.rm = TRUE
  ) +
  scale_y_continuous(
    limits = c(0, 50),
    labels = scales::label_percent(scale = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "Surveillance Year",
    y = "Positivity Rate (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line.x = element_line(color = "grey60", linewidth = 0.4)
  )
```


```{r}
#| label: fig-toronto-ticks-map
#| fig-cap: "Blacklegged ticks cluster in the east end of Toronto, around the Rouge Valley, with a smaller cluster on the Toronto Islands. Each circle is one surveillance site, with counts summed over all years surveyed. Hollow grey circles mark sites where no ticks were ever found; red circles are sized by the total number of adult and nymph ticks collected. The grey outline is the City of Toronto boundary."
#| echo: false
#| fig-width: 7
#| fig-height: 4.5

# Read the local geographic boundary file
toronto_boundary <- read_sf(here("data", "01-raw_data", "toronto_boundary.gpkg"))

# Aggregate total ticks found per site across all surveillance years
park_summary <- analysis_data |>
  group_by(park_locations) |>
  summarize(
    total_ticks = sum(total_blts, na.rm = TRUE),
    latitude = first(latitude),
    longitude = first(longitude),
    .groups = "drop"
  ) |>
  # Draw small circles last so they sit on top of large ones
  arrange(desc(total_ticks))

# Split data into zero-count sites and positive-count sites
zero_parks <- park_summary |> filter(total_ticks == 0)
pos_parks  <- park_summary |> filter(total_ticks > 0)

# Plot geographic map
ggplot() +
  # Base map of Toronto boundary
  geom_sf(
    data = toronto_boundary,
    fill = "grey97",
    color = "grey60",
    linewidth = 0.3
  ) +
  # Small hollow grey circles for sites with 0 ticks
  geom_point(
    data = zero_parks,
    aes(x = longitude, y = latitude),
    shape = 1,          # Hollow circle
    color = "grey55",
    size = 2.5,
    stroke = 0.8,
    alpha = 0.6
  ) +
  # Sized, red circles with a thin white outline for detected ticks
  geom_point(
    data = pos_parks,
    aes(x = longitude, y = latitude, size = total_ticks),
    shape = 21,
    fill = "#8b0000",
    color = "white",
    stroke = 0.6,
    alpha = 0.75
  ) +
  # Geographic text labels
  annotate("text", x = -79.18, y = 43.838, label = "Rouge Valley", size = 3.5, fontface = "italic", color = "grey20") +
  annotate("text", x = -79.38, y = 43.603, label = "Toronto Islands", size = 3.5, fontface = "italic", color = "grey20") +
  annotate("text", x = -79.38, y = 43.655, label = "Downtown", size = 3.5, fontface = "bold", color = "grey20") +
  annotate("text", x = -79.45, y = 43.585, label = "Lake Ontario", size = 4.5, fontface = "italic", color = "steelblue") +
  # Scales and Coordinate system
  scale_size_continuous(
    breaks = c(10, 50, 200),
    range = c(2.5, 9),
    name = "Ticks Collected (all years)"
  ) +
  coord_sf(
    xlim = c(-79.65, -79.10),
    ylim = c(43.57, 43.85), # Expanded slightly to fit Lake Ontario label
    expand = FALSE
  ) +
  theme_void(base_size = 12) +
  theme(
    legend.position = "right"
  )
```

```{r}
#| label: fig-top10-parks
#| fig-cap: "Top 10 Toronto parks with the highest cumulative blacklegged tick counts across all surveillance years."
#| echo: false
#| fig-height: 5
#| warning: false
#| message: false

# Aggregate ticks per park, filter for top 10, and sort ascending for horizontal display
top10_parks <- analysis_data |> 
  group_by(park_locations) |> 
  summarize(
    total_ticks = sum(total_blts, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  filter(total_ticks >= 1) |> 
  slice_max(order_by = total_ticks, n = 10, with_ties=FALSE) |> 
  mutate(
    park_locations = fct_reorder(park_locations, total_ticks)
  )

# Horizontal bar chart
ggplot(top10_parks, aes(x = total_ticks, y = park_locations)) +
  geom_col(
    fill = "#8b0000",
    width = 0.65
  ) +
  geom_text(
    aes(label = scales::comma(total_ticks)),
    hjust = -0.2,
    size = 3.5,
    color = "grey20"
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.12)),
    labels = scales::comma
  ) +
  scale_y_discrete(labels = scales::label_wrap(24)) +
  labs(
    x = "Total Blacklegged Ticks (Adults & Nymphs)",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.line.x = element_line(color = "grey60", linewidth = 0.4),
    axis.text.y = element_text(size = 10, face = "bold", color = "grey20"),
  )
```

# References 


