---
title: "My title"
subtitle: "My subtitle if needed"
author: "Shrey Sati"
thanks: "Code and data are available at: [https://github.com/UpCatch/Toronto_Tick_Surveillance_Data](https://github.com/UpCatch/Toronto_Tick_Surveillance_Data)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(tinytable)
```


# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....






# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@blacklegged_tick_data].... Following @tellingstories, we consider...

Overview text

## Measurement
	
Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

```{r}
#| label: fig-tick-surveillance
#| fig-cap: "Distribution of total blacklegged ticks collected per park across surveillance years in Toronto."
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(here)

# Load the dataset
analysis_data <- read_csv(here("data", "02-analysis_data", "analysis_data.csv"))

# Convert year to factor with full levels from 2013 to 2023 to display gap years
full_years <- as.character(2013:2023)

plot_data <- analysis_data |> 
  mutate(year = factor(as.character(year), levels = full_years))

# Plot raw points with horizontal jitter
set.seed(853)

ggplot(plot_data, aes(x = year, y = total_blts)) +
  geom_jitter(
    width = 0.35,
    height = 0,
    alpha = 0.4,
    size = 1.5,
    color = "darkred"
  ) +
  scale_x_discrete(drop = FALSE) +
  labs(
    x = "Surveillance Year",
    y = "Total Blacklegged Ticks (BLTs)",
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank()
  )
```

```{r}
#| label: fig-annual-total-ticks
#| fig-cap: "Total adult and nymph blacklegged ticks collected per year across all surveyed parks in Toronto."
#| echo: false
#| warning: false
#| message: false

# Aggregate total ticks by year and retain all years 2013-2023
annual_ticks <- analysis_data |> 
  group_by(year) |> 
  summarize(sum_blts = sum(total_blts, na.rm = TRUE), .groups = "drop") |> 
  complete(
    year = 2013:2023,
    fill = list(sum_blts = NA)
  ) |> 
  mutate(year = factor(year, levels = 2013:2023))

# Plot bar chart
ggplot(annual_ticks, aes(x = year, y = sum_blts)) +
  geom_col(
    fill = "#2c7bb6",
    width = 0.65,
    na.rm = TRUE
  ) +
  geom_text(
    aes(label = ifelse(is.na(sum_blts), "", scales::comma(sum_blts))),
    vjust = -0.5,
    size = 3.3,
    color = "grey20"
  ) +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.12)),
    labels = scales::comma
  ) +
  labs(
    x = "Year",
    y = "Total Blacklegged Ticks (Adults & Nymphs)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line.x = element_line(color = "grey60", linewidth = 0.4),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey35", margin = margin(b = 10))
  )
```


```{r}
#| label: fig-positivity-rate
#| fig-cap: "Annual blacklegged tick positivity rate across Toronto parks (2015–2023), with the number of positive ticks annotated."
#| echo: false
#| warning: false
#| message: false

# Calculate annual positivity rates and complete timeline for 2015-2023
positivity_data <- analysis_data |> 
  filter(year >= 2015) |> 
  group_by(year) |> 
  summarize(
    total_positive = sum(num_positive, na.rm = TRUE),
    total_tested = sum(blt_adults_and_nymphs, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    pct_positive = (total_positive / total_tested) * 100
  ) |> 
  complete(
    year = 2015:2023,
    fill = list(pct_positive = NA, total_positive = NA)
  ) |> 
  mutate(
    year_factor = factor(year, levels = 2015:2023)
  )

# Plot line chart
ggplot(positivity_data, aes(x = year_factor, y = pct_positive, group = 1)) +
  # Shade the years with no surveillance (x positions 6 to 8 are 2020 to 2022)
  #annotate("rect", xmin = 5.5, xmax = 8.5, ymin = -Inf, ymax = Inf, fill = "grey90") +
  #annotate("text", x = 7, y = 25, label = "No surveillance", colour = "grey40") +
  geom_line(
    color = "#8B0000",
    linewidth = 0.9,
    na.rm = TRUE
  ) +
  geom_point(
    color = "#8B0000",
    size = 2.8,
    na.rm = TRUE
  ) +
  geom_text(
    aes(label = ifelse(is.na(total_tested), "", paste0("n = ", total_tested))),
    vjust = -1.2,
    size = 3.2,
    color = "grey25",
    na.rm = TRUE
  ) +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(
    limits = c(0, 50),
    labels = scales::label_percent(scale = 1),
    expand = expansion(mult = c(0.05, 0.18))
  ) +
  labs(
    x = "Surveillance Year",
    y = "Positivity Rate (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line.x = element_line(color = "grey60", linewidth = 0.4)
  )
```

```{r}
#| label: fig-toronto-ticks-map
#| fig-cap: "Geographic distribution of blacklegged tick surveillance sites in Toronto. Each dot represents a park summary aggregated over all surveillance years."
#| echo: false
#| warning: false
#| message: false

# Aggregate total ticks found per site across all surveillance years
park_summary <- analysis_data |> 
  group_by(park_locations, latitude, longitude) |> 
  summarize(
    total_ticks = sum(total_blts, na.rm = TRUE),
    .groups = "drop"
  )

# Split data into zero-count sites and positive-count sites
zero_parks <- park_summary |> filter(total_ticks == 0)
pos_parks  <- park_summary |> filter(total_ticks > 0)

# Plot geographic scatter
ggplot() +
  # Hollow grey circles for sites with 0 ticks
  geom_point(
    data = zero_parks,
    aes(x = longitude, y = latitude),
    shape = 1,          # Hollow circle
    color = "grey55",
    size = 2.5,
    stroke = 0.8,
    alpha = 0.6
  ) +
  # Sized, dark red circles for sites where ticks were detected
  geom_point(
    data = pos_parks,
    aes(x = longitude, y = latitude, size = total_ticks),
    shape = 16,         # Solid circle
    color = "#8b0000",  # Dark red
    alpha = 0.55
  ) +
  scale_size_continuous(
    limits = c(1, 150),
    breaks = c(1, 25, 50, 100, 150),
    range = c(2.5, 9),
    name = "Ticks Detected"
  ) +
  coord_quickmap(
    xlim = c(-79.62, -79.14),
    ylim = c(43.59, 43.83)
  ) +
  labs(
    x = "Longitude (°W)",
    y = "Latitude (°N)",
    title = "Blacklegged Tick Surveillance Sites Across Toronto Parks",
    subtitle = "Hollow grey circles denote zero detections; red circles indicate tick presence",
    caption = "Note: Each dot is a summary over all years."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#fafafa", color = NA),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey35", margin = margin(b = 10)),
    plot.caption = element_text(face = "italic", color = "grey40", size = 9, margin = margin(t = 10)),
    legend.position = "right"
  )
```

```{r}
#| label: fig-top10-parks
#| fig-cap: "Top 10 Toronto parks with the highest cumulative blacklegged tick counts across all surveillance years."
#| fig-height: 5
#| warning: false
#| message: false

# Aggregate ticks per park, filter for top 10, and sort ascending for horizontal display
top10_parks <- analysis_data |> 
  group_by(park_locations) |> 
  summarize(
    total_ticks = sum(total_blts, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  filter(total_ticks >= 1) |> 
  slice_max(order_by = total_ticks, n = 10) |> 
  mutate(
    park_locations = fct_reorder(park_locations, total_ticks)
  )

# Horizontal bar chart
ggplot(top10_parks, aes(x = total_ticks, y = park_locations)) +
  geom_col(
    fill = "#8b0000",
    width = 0.65
  ) +
  geom_text(
    aes(label = scales::comma(total_ticks)),
    hjust = -0.2,
    size = 3.5,
    color = "grey20"
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.12)),
    labels = scales::comma
  ) +
  labs(
    x = "Total Blacklegged Ticks (Adults & Nymphs)",
    y = NULL,
    title = "Top 10 Toronto Parks by Blacklegged Tick Counts",
    subtitle = "Aggregated across all surveillance years (2013–2019, 2023)",
    caption = "Data: Toronto Public Health Blacklegged Tick Surveillance"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.line.x = element_line(color = "grey60", linewidth = 0.4),
    axis.text.y = element_text(size = 10, face = "bold", color = "grey20"),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey35", margin = margin(b = 10)),
    plot.caption = element_text(color = "grey40", size = 8, margin = margin(t = 10))
  )
```

# References 


