---
title: "Blacklegged Ticks in Toronto Parks Are Concentrated in the East"
subtitle: "Three-quarters of surveys from 2013 to 2023 found no ticks, while five Rouge Valley sites produced two-thirds of the ticks collected and most of the infected ones"
author: "Shrey Sati"
thanks: "Code and data are available at: [https://github.com/UpCatch/Toronto_Tick_Surveillance_Data](https://github.com/UpCatch/Toronto_Tick_Surveillance_Data)."
date: today
date-format: long
abstract: "We analyze Toronto Public Health's surveillance of blacklegged ticks, which carry the bacterium that causes Lyme disease, using 233 surveys of 84 sites in city parks from 2013 to 2023. Three-quarters of surveys found no ticks, while five Rouge Valley sites, in the city's northeast, produced 65% of the 606 ticks collected and 83% of those that tested positive. The rise in infected ticks after 2015 reflects how much of each year's catch came from the Rouge Valley, not a change across Toronto. Knowing that Lyme disease risk is concentrated in a handful of parks can help focus prevention where it is needed most."
format: pdf
number-sections: true
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false

library(tidyverse)
library(here)
library(sf)
library(tinytable)

# Load the dataset
analysis_data <- read_csv(here("data", "02-analysis_data", "analysis_data.csv"))
```


# Introduction

As summers get warmer and the threat of climate change looms, Canadians have yet
another thing to worry about: Lyme disease. But how big an issue is it, and is
climate change actually linked to it? Lyme disease is becoming more common in
Canada, with reported cases rising from 338 in 2012 to 2,525 in 2022 [@wafy2026]
and preliminary counts passing 5,200 in 2024 [@dev2026]. The disease is spread by
the blacklegged tick, also known as the deer tick, which carries the bacterium that
causes it. Over recent decades, the blacklegged tick's range and numbers have
continued to grow [@eisen2016], and Ontario is at the centre of this change:
in the first half of 2026, the province accounted for nearly two-thirds of blacklegged tick reports
submitted by the public in Canada [@otis2026]. All of Toronto is now considered
an estimated risk area for Lyme disease [@tph_lyme]. However, national surveillance
reports only by province, and public reports depend on who notices a tick, if
anyone does, so neither can say where within a city the risk lies.

In this paper, we analyze all 233 surveys in Toronto Public Health's Blacklegged Tick Surveillance dataset, covering 84 sites between 2013 and 2023. Three-quarters of surveys found no ticks, and 52 of the 84 sites never recorded one. Most ticks were concentrated in a few places in the east of the city. Five sites in the Rouge Valley produced 65% of the 606 ticks collected, and a smaller cluster appeared on the Toronto Islands. Infected ticks were even more concentrated. 45% of Rouge Valley ticks tested positive, compared with 17% elsewhere, and these five sites held 83% of all positive ticks. The citywide share of positive ticks rose after 2015, but this reflects how much of each year's catch came from the Rouge Valley rather than a trend across the city, and the data cannot show whether climate change played a role. This matters because surveillance is meant to show where and when to focus prevention [@wafy2026]. It also shapes how the data should be read: the program samples in the fall and so largely misses nymphs, the tick stage most likely to infect people [@eisen2016].

The remainder of this paper is structured as follows. @sec-data describes the data source, how ticks are collected and tested, each variable in the dataset, and the patterns in tick counts, locations and infection. @sec-appendix-cleaning describes how the raw data were cleaned and tested.

# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR]. The dataset used in this paper comes from Toronto Public Health's
Blacklegged Tick Surveillance dataset [@blacklegged_tick_data], which was obtained
from the City of Toronto's Open Data Portal [@opendatatoronto]. Additional packages in R 
that helped to support this paper include `tidyverse` [@tidyverse], `ggplot2` [@ggplot2]
and `tinytable` [@tinytable] for graphs and tables respectively, `here` [@here] for
handling file paths, and the map for the City of Toronto's municipal boundary [@toronto_boundary],
which `sf` [@sf] helped to draw. With inspiration from Rohan Alexander's book [@tellingstories],
we focus on what the data can tell us about the world we inhabit.

In Ontario, blacklegged ticks (*Ixodes scapularis*) are the only ticks that transmit
*Borrelia burgdorferi*, the bacterium that causes Lyme disease [@tph_lyme]. Over
the last two decades, Lyme disease in Canada and in the United States has increased 
both in incidence and geographic extent [@eisen2016]. Eisen et al. explain that a tick's development, survival,
and its ability to find a host are directly affected by temperature and humidity. 
Compared to a generalized life cycle of approximately 2 years for ticks inhabiting
the northeastern United States, the life cycle of some ticks in comparatively
colder southeastern Canada may take 3 to 4 years, which lowers their reproductive
rate [@eisen2016]. Justin Wood, founder of the tick-testing company Geneticks, told CP24 that climate
change is driving the rise in tick reports. Ontario accounted for 63.7% of
blacklegged tick reports across Canada in the first half of 2026, and the province's
reports rose 72.8% from the same period in 2025 [@otis2026]. Toronto is now
considered to be an Estimated Risk Area for Lyme disease, but Toronto Public
Health considers the risk of infection low, although increasing [@tph_lyme]. National reports 
note that regular surveillance activity helps identify any shifting trends
in Lyme disease occurrence and geographic distribution [@wafy2026].

Every row in the dataset represents a single site surveyed in a year. In total,
there are 233 rows detailing 84 site names covered from 2013 to 2019 and 2023.
No surveillance was conducted from 2020 to 2022. For each site and year in the dataset,
Toronto Public Health records:

- **Site name** and **location** (as a single latitude and longitude representing
the entire site)
- **Year** the surveillance took place
- **Adult and nymph ticks collected** (the last two stages in a blacklegged 
tick's life cycle)
- **Larvae collected** (newly hatched ticks, the first stage in the life cycle)
- **Ticks that tested positive** (assumed to be for *Borrelia burgdorferi*; see below)

There is another column in the dataset named "Total BLTs", but given that it is
identical to the adult and nymph counts in all rows, we treat the two columns
as one variable. Furthermore, the **Rouge Valley**, located on the Northeastern side of
Toronto, contains 5 sites including the name "Rouge". To learn more about the
data cleaning process for this dataset, such as how the latitudes and longitudes
were extracted, and how sites with similar names were handled, please see
@sec-appendix-cleaning.

## Measurement
	
Toronto Public Health's method of surveillance is a process termed "tick dragging",
involving a surveyor dragging a white flannel cloth behind them across the ground
and through vegetation, waiting for ticks to latch on [@pho_tick_procedure]. Sites
are chosen according to how suitable they are for blacklegged ticks to inhabit or
if there was a previously confirmed finding, and drags are conducted during the
fall when adult ticks are at their most active [@tph_surveillance]. Collected ticks
are removed, placed into a collection vial, and shipped to the National Microbiology
Laboratory to identify species and run diagnostics [@pho_tick_procedure]. The
results for every site then make up one row for that particular year.

While the dataset does not explicitly name the pathogen identified in the positive
count column, Toronto Public Health states that testing is done for *B. burgdorferi*,
and as of 2023, for the *Babesia* species, *Anaplasma phagocytophilum* and the
Powassan virus. Given that the surveillance program started with the goal of
tracking blacklegged ticks carrying Lyme disease as its core focus, we assume that
the positive count values represent ticks with the *B. burgdorferi* bacteria present.

Each step of the method shapes what the data mean. Firstly, the count of blacklegged ticks
at a given site is a sample rather than a census. Tick dragging fails to account
for ticks attached to animals, and a drag may not "catch" every single tick along
the path. The number of ticks found can also depend on the weather or the care
of the surveyor. Toronto Public Health notes that there may be ticks present at a
site even if none were reportedly collected [@tph_surveillance]. Secondly, the sites
are chosen based on past tick findings or their suitability for ticks to inhabit.
This choice, while understandable, can have the effect of overrepresenting sites
with ticks. Lastly, dragging in the fall misses the peak of nymph activity, which runs from
late May to early July [@eisen2016]. Dragging is timed for adults partly because
nymphs are so small [@pho_tick_procedure], but nymphs matter most for people: they
are more likely than adults to stay attached long enough to transmit infection
[@eisen2016].  

## Ethical considerations

The primary ethical consideration is that naming parks where ticks were found 
might prevent people from visiting them. This is possibly made worse by the fact
that most sites with large tick presence are in Scarborough, in Toronto's east end. Furthermore, a site not recording any
ticks does not confirm that no ticks are present [@tph_surveillance]. The dataset
may be interpreted by the public as a ranking of which sites are safe to visit and
which are not, when it should be perceived as which sites require the greatest amount
of precaution.

## Similar datasets

The Public Health Agency of Canada publishes a national summary once annually that
combines tick counts from dragging with images of ticks submitted by the public,
primarily through the online, image-based platform eTick.ca [@wafy2026]. Otis notes
that because eTick relies on users to submit data, confirmed numbers likely
underestimate actual population numbers in Canada [@otis2026]. Accurate numbers
from public submissions also rely on individuals noticing a tick and choosing to
report it, which is not always the case. Toronto Public Health's dataset was
the only source found that records systematic tick collection at named locations in the city.

## Summary of the variables

@tbl-variable-summary takes the three count variables and provides a summary
across all site visits for every year surveillance occurred. The mean represents
the average tick tally per survey, the median is the middle value when surveys are
ordered by count in ascending order, and the standard deviation shows the degree
to which the counts are spread around the mean. The median is 0 for all variables,
and the standard deviation is considerably higher than the mean in all cases, which
means most surveys found nothing while a few found many ticks.

```{r}
#| label: tbl-variable-summary
#| tbl-cap: "Summary statistics for tick counts across all 233 surveys (one per site per year)."
#| echo: false
#| warning: false
#| message: false

# Summary statistics for the three count variables across all 233 surveys
summary_df <- data.frame(
  Variable = c("Adults and nymphs", "Larvae", "Tested positive"),
  Mean = c(
    mean(analysis_data$blt_adults_and_nymphs),
    mean(analysis_data$blt_larvae),
    mean(analysis_data$num_positive)
  ),
  Median = c(
    median(analysis_data$blt_adults_and_nymphs),
    median(analysis_data$blt_larvae),
    median(analysis_data$num_positive)
  ),
  `Standard deviation` = c(
    sd(analysis_data$blt_adults_and_nymphs),
    sd(analysis_data$blt_larvae),
    sd(analysis_data$num_positive)
  ),
  Maximum = c(
    max(analysis_data$blt_adults_and_nymphs),
    max(analysis_data$blt_larvae),
    max(analysis_data$num_positive)
  ),
  Total = c(
    sum(analysis_data$blt_adults_and_nymphs),
    sum(analysis_data$blt_larvae),
    sum(analysis_data$num_positive)
  ),
  # Number of surveys with at least one tick in each category
  `Surveys above zero` = c(
    sum(analysis_data$blt_adults_and_nymphs > 0),
    sum(analysis_data$blt_larvae > 0),
    sum(analysis_data$num_positive > 0)
  ),
  check.names = FALSE
)

summary_df |>
  tt() |>
  # Mean and standard deviation to exactly two decimal places
  format_tt(j = c(2, 4), digits = 2, num_fmt = "decimal", num_zero = TRUE) |>
  # Whole-number columns with thousands separators
  format_tt(j = c(3, 5, 6, 7), digits = 0, num_mark_big = ",") |>
  style_tt(j = 2:7, align = "r")
```


@tbl-annual-tick-summary showcases the same counts broken down by year. From the
table, we can see that the number of sites surveyed went from 1 in 2013 to 36 in
2016. This suggests that a year with a greater tick count can be a result of the
increase in sites searched. Thus, yearly totals are not a good indicator of the
tick count in Toronto.

```{r}
#| label: tbl-annual-tick-summary
#| tbl-cap: "Sites surveyed, ticks collected and ticks that tested positive by year, 2013–2023."
#| echo: false
#| warning: false
#| message: false

# Summarize the data by year
summary_table <- analysis_data |>
  group_by(year) |>
  summarize(
    `Parks surveyed` = n(),
    `Parks with ticks` = sum(total_blts > 0, na.rm = TRUE),
    `Adults and nymphs` = sum(blt_adults_and_nymphs, na.rm = TRUE),
    `Larvae` = sum(blt_larvae, na.rm = TRUE),
    `Tested positive` = sum(num_positive, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    `Positive (%)` = (`Tested positive` / `Adults and nymphs`) * 100
  ) |>
  rename(Year = year)

# Generate the tinytable with formatting
tt(summary_table, width = 1) |>
  format_tt(j = 2:6, digits = 0, num_mark_big = ",") |>
  # num_fmt = "decimal" gives one decimal place (the default rounds to significant figures)
  format_tt(j = 7, digits = 1, num_fmt = "decimal", num_zero = TRUE) |>
  # Escape the % in the column name so LaTeX prints it instead of treating it as a comment
  format_tt(escape = TRUE) |>
  style_tt(j = 2:7, align = "r")
```

## Most surveys found no ticks
Looking at all 233 surveys, 176 (76%) of them found no blacklegged ticks
(@fig-tick-surveillance). While the average survey found 2.6 ticks, this is inflated
by some sites having much larger counts (e.g., Rouge Park in 2017 with 122 ticks found).
Over all surveys, 606 adults and nymphs were collected.

```{r}
#| label: fig-tick-surveillance
#| fig-cap: "Adult and nymph ticks collected in each survey, 2013–2023; each dot is one site in one year."
#| echo: false
#| warning: false
#| message: false

# Convert year to factor with full levels from 2013 to 2023 to display gap years
full_years <- as.character(2013:2023)

plot_data <- analysis_data |>
  mutate(year = factor(as.character(year), levels = full_years))

# Plot raw points with horizontal jitter
set.seed(853)

ggplot(plot_data, aes(x = year, y = total_blts)) +
  geom_jitter(
    width = 0.35,
    height = 0,
    alpha = 0.4,
    size = 1.5,
    color = "darkred"
  ) +
  scale_x_discrete(drop = FALSE) +
  labs(
    x = "Surveillance Year",
    y = "Total Blacklegged Ticks (BLTs)",
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank()
  )
```

@tbl-annual-tick-summary shows that larvae were rare, with just 22 being found in total. This
tally comes from 4 surveys at 3 different sites (Anewen Greenbelt in 2017 and 2018,
Highland Creek Park in 2017, and Cedar Ridge Park in 2018). Larvae hatch from eggs
laid by fed adult females [@eisen2016], which could suggest that ticks were reproducing
at those sites. Larvae are not tested and are not part of the total tick count,
so we exclude them from the figures that follow.

## Ticks are concentrated in the east

The majority of ticks were on the east side of Toronto (@fig-toronto-ticks-map).
Sites that were east of longitude 79.25°W produced around 88% of all collected ticks.
The Rouge Valley sites (five in total) produced 65% of the total ticks found
(392 ticks). There are smaller clusters around the Toronto Islands, but most sites
never recorded any ticks. Of all 84 sites surveyed, 52 had no ticks in any year.

```{r}
#| label: fig-toronto-ticks-map
#| fig-cap: "Surveillance sites across Toronto, sized by total ticks collected; hollow grey circles mark sites where no ticks were found."
#| echo: false
#| fig-width: 7
#| fig-height: 4.5

# Read the local geographic boundary file
toronto_boundary <- read_sf(here("data", "01-raw_data", "toronto_boundary.gpkg"))

# Aggregate total ticks found per site across all surveillance years
park_summary <- analysis_data |>
  group_by(park_locations) |>
  summarize(
    total_ticks = sum(total_blts, na.rm = TRUE),
    latitude = first(latitude),
    longitude = first(longitude),
    .groups = "drop"
  ) |>
  # Draw small circles last so they sit on top of large ones
  arrange(desc(total_ticks))

# Split data into zero-count sites and positive-count sites
zero_parks <- park_summary |> filter(total_ticks == 0)
pos_parks <- park_summary |> filter(total_ticks > 0)

# Plot geographic map
ggplot() +
  # Base map of Toronto boundary
  geom_sf(
    data = toronto_boundary,
    fill = "grey97",
    color = "grey60",
    linewidth = 0.3
  ) +
  # Small hollow grey circles for sites with 0 ticks
  geom_point(
    data = zero_parks,
    aes(x = longitude, y = latitude),
    shape = 1, # Hollow circle
    color = "grey55",
    size = 2.5,
    stroke = 0.8,
    alpha = 0.6
  ) +
  # Sized, red circles with a thin white outline for detected ticks
  geom_point(
    data = pos_parks,
    aes(x = longitude, y = latitude, size = total_ticks),
    shape = 21,
    fill = "#8b0000",
    color = "white",
    stroke = 0.6,
    alpha = 0.75
  ) +
  # Geographic text labels
  annotate("text", x = -79.18, y = 43.838, label = "Rouge Valley", size = 3.5, fontface = "italic", color = "grey20") +
  annotate("text", x = -79.38, y = 43.603, label = "Toronto Islands", size = 3.5, fontface = "italic", color = "grey20") +
  annotate("text", x = -79.38, y = 43.655, label = "Downtown", size = 3.5, fontface = "bold", color = "grey20") +
  annotate("text", x = -79.45, y = 43.585, label = "Lake Ontario", size = 4.5, fontface = "italic", color = "steelblue") +
  # Scales and Coordinate system
  scale_size_continuous(
    breaks = c(10, 50, 200),
    range = c(2.5, 9),
    name = "Ticks Collected (all years)"
  ) +
  coord_sf(
    xlim = c(-79.65, -79.10),
    ylim = c(43.57, 43.85), # Expanded slightly to fit Lake Ontario label
    expand = FALSE
  ) +
  theme_void(base_size = 12) +
  theme(
    legend.position = "right"
  )
```


@fig-top10-parks showcases the top ten sites with the greatest number of ticks.
Rouge Park recorded 209, and the Rouge Valley sites occupy five spots in the top
10, with Colonel Danforth Trail and Morningside Park also representing Toronto's
east end. This illustrates the pattern that climate assigns a broad area
where ticks can inhabit, but actual tick numbers within that region depend on
woodland habitat and host densities [@eisen2016]. Rouge Valley is a large woodland
area, the type of habitat that would be expected to hold more ticks.

```{r}
#| label: fig-top10-parks
#| fig-cap: "The ten sites with the most blacklegged ticks, summed across all surveillance years."
#| echo: false
#| fig-height: 5
#| warning: false
#| message: false

# Aggregate ticks per park, filter for top 10, and sort ascending for horizontal display
top10_parks <- analysis_data |>
  group_by(park_locations) |>
  summarize(
    total_ticks = sum(total_blts, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(total_ticks >= 1) |>
  slice_max(order_by = total_ticks, n = 10, with_ties = FALSE) |>
  mutate(
    park_locations = fct_reorder(park_locations, total_ticks)
  )

# Horizontal bar chart
ggplot(top10_parks, aes(x = total_ticks, y = park_locations)) +
  geom_col(
    fill = "#8b0000",
    width = 0.65
  ) +
  geom_text(
    aes(label = scales::comma(total_ticks)),
    hjust = -0.2,
    size = 3.5,
    color = "grey20"
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.12)),
    labels = scales::comma
  ) +
  scale_y_discrete(labels = scales::label_wrap(24)) +
  labs(
    x = "Total Blacklegged Ticks (Adults & Nymphs)",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.line.x = element_line(color = "grey60", linewidth = 0.4),
    axis.text.y = element_text(size = 10, face = "bold", color = "grey20")
  )
```

## Infected ticks are concentrated in large catches

Of all 606 adults and nymphs collected, 35% or 211 ticks tested positive for the
*Borrelia burgdorferi* bacterium, compared with 27.4% collected by dragging across
Ontario in 2022 [@wafy2026]. Most infected ticks were present in the small number
of large catches (@fig-positive-vs-collected). The 13 surveys that found 10 or
more ticks accounted for 82% of all ticks found but 96% of positive ones. 

```{r}
#| label: fig-positive-vs-collected
#| fig-cap: "Ticks that tested positive against ticks collected, for the 57 surveys that found at least one tick."
#| fig-width: 7
#| fig-height: 4.5
#| echo: false
#| warning: false
#| message: false

# Keep surveys that collected at least one adult or nymph, and flag Rouge Valley sites
positive_data <- analysis_data |>
  filter(blt_adults_and_nymphs > 0) |>
  mutate(
    area = if_else(
      str_detect(park_locations, "Rouge"),
      "Rouge Valley sites",
      "Other sites"
    )
  )

# pseudo_log stretches small values but, unlike a log scale, can show zero
pseudo_log_axis <- scales::pseudo_log_trans(base = 10)

set.seed(853)

ggplot(positive_data, aes(x = blt_adults_and_nymphs, y = num_positive, colour = area)) +
  # Reference line: every tick collected tested positive
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(width = 0.03, height = 0.03, size = 2.5, alpha = 0.8) +
  scale_x_continuous(trans = pseudo_log_axis, breaks = c(1, 3, 10, 30, 100)) +
  scale_y_continuous(trans = pseudo_log_axis, breaks = c(0, 1, 3, 10, 30, 60)) +
  scale_colour_manual(values = c("Rouge Valley sites" = "#8b0000", "Other sites" = "grey50")) +
  labs(
    x = "Adult and nymph ticks collected",
    y = "Ticks that tested positive",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

The positivity percentage increased sharply from just under 6% in 2015 to between
29% and 43% in the years that followed (@tbl-annual-tick-summary), but we cannot
attribute this to climate change. This jump is largely from Rouge Park, where
29 of 63 ticks tested positive in 2016 alone. Outside the Rouge Valley, the data
show no steady trend in this regard. This reflects the literature: the spread of blacklegged ticks in Canada has
coincided with warming, but no change in exposure has yet been conclusively linked
to climate change [@eisen2016].

\newpage

# References

::: {#refs}
:::

\newpage

\appendix

# Data cleaning {#sec-appendix-cleaning}

The raw data were downloaded from the Open Data Portal with `opendatatoronto` [@opendatatoronto], which was then saved before cleaning. Cleaning made four changes. Each site's location is stored in the raw data as a text field in GeoJSON format. The first change was to extract the longitude and latitude from this text into two numeric columns. Second, the columns were renamed to consistent, readable names, and the original location field was dropped. Third, the count columns and year were converted to whole numbers. Fourth, only the columns used in this paper were kept. There were no rows that were removed.

Two issues were noted but not changed, so that the data match what Toronto Public Health published. Two sites, Upper Rouge Trail Park and Cedar Ridge Park, have slightly different coordinates in different years. The cleaned data keep both, and only the map in @fig-toronto-ticks-map places each site at its first recorded location so that it appears once. There are also several sites that appear under more than one name, such as "Wards Island" and "Toronto Islands: Ward's", or "Guild Park" and "Guild Park and Gardens". These were left as recorded, which is why we refer to 84 site names rather than 84 distinct places.

The cleaned data were checked with tests written using `testthat` [@testthat], which confirm the number of rows and columns, that each year is a surveillance year, that site names are never missing, that the total count equals the adult and nymph count, that positive ticks never exceed ticks collected, and that every location falls within Toronto.
